The dynamics of starvation and recovery

### 2016-07-29 Manuscript/4.tex
Kempes edits

\noindent {\bf Nutritional-state-structured model (NSM)}\\
We begin by defining a minimal Nutritional-State-structured population Model
(NSM), where the consumer population is divided into two energetic states:
(a) an energetically replete (full) state $F$, where the consumer reproduces
at a constant rate $\lambda$ and has no mortality risk, and (b) an
energetically deficient (hungry) state $H$, where the consumer does not
reproduce but dies at rate $\mu$.  The underlying resource $R$ evolves by
logistic growth with an intrinsic growth rate $\alpha$ and a carrying
capacity equal to one.  Consumers transition from the full state $F$ to the
hungry state $H$ by starvation at rate $\sigma$ and also in proportion to the
absence of resources $(1-R)$.  Conversely, consumers recover from state $H$
to state $F$ at rate $\rho$ and in proportion to $R$.  Resources are also
eaten by the consumers---at rate $\rho$ by hungry consumers and at rate
$\beta<\rho$ by full consumers.  This inequality accounts for hungry
consumers requiring more resources to rebuild body weight.

@@

Nearly all of the rates described in the NSM are to some extent governed by
consumer metabolism, which can be used to describe a variety of organismal
features~\cite{}\sid{Ref}.  The scaling relation between an organism's
metabolic rate $B$ and its body size at reproductive maturity $M$ is well
documented~\cite{West:2002it} and scales as $B = B_0 M^\eta$, where $\eta$ is
the scaling exponent, generally assumed to vary around $2/3$ or $3/4$ for
metazoans~\cite{}\sid{Ref}, and has taxonomic shifts for unicellular species
between $\eta\approx 1$ in eukaryotes and $\eta\approx 1.76$ in bacteria
\cite{DeLong:2010dy,Kempes:2012hy}. Several efforts have shown how a
partitioning of this metabolic rate between growth and maintenance purposes
can be used to derive a general equation for the growth trajectories and
growth rates of organisms ranging from bacteria to metazoans
\cite{Kempes:2012hy,more}\sid{fix Ref}. More specifically, the interspecific
\sid{what does interspecific mean?}  trends in growth rate can be
approximated by
\begin{eqnarray}
\lambda = \lambda_0 M^{\eta-1}.
\label{lambda}
\end{eqnarray}
This relation is derived from the simple balance condition
\begin{eqnarray}
\label{balance}
B_{0}m^{\eta}=E_{m}\frac{dm}{dt}+B_{m}m\,,
\end{eqnarray}
where $E_{m}$ is the energy needed to synthesize a unit of mass, $B_{m}$ is
the metabolic rate to support an existing unit of mass, and $m$ is the mass
at any point in development. It is useful to explicitly write this balance
because it can also be modified to understand the timescales of both
starvation and recovery from starvation as we show below.

To determine the starvation rate, $\sigma$, we are interested in the time
required for an organism to go from a mature adult that reproduces at rate
$\lambda$ (henceforth we term this state as the ``replete'' state), to a
reduced-size hungry state where reproduction is impossible.  This transition
time can be inferred from the energy balance in Eq.~\ref{balance}, where we
make the basic assumption that an organism must meet its maintenance
requirements using the digestion of existing mass as the sole energy source.
This assumption implies the simple metabolic balance \sid{don't understand
  this equation}
\begin{eqnarray}
\frac{dm}{dt}E_{m}^{\prime}=-B_{m}m
\end{eqnarray}
where $E_{m}^{\prime}$ is the amount of energy stored in a unit of existing
body mass which differs from $E_{m}$~\cite{}\sid{Ref}, the energy required to
synthesis a unit of biomass. Given the replete mass, $M$, of an organism, the
above energy balance prescribes the mass trajectory of a non-consuming
organism:
\begin{eqnarray}
\label{mt}
m\left(t\right)=Me^{-B_{m}t/E_{m}^{\prime}}.
\end{eqnarray}
Since only certain tissues can be digested for energy (for example the brain
cannot be degraded to fuel metabolism), we define the rate for starvation and
death by the timescales required to reach specific fractions of the
replete-state mass.  We define $m_{\rm starve}=\epsilon M$, where
$\epsilon<1$ is the fraction of replete-state mass where reproduction ceases.
This fraction will be modified if tissue composition systematically scales
with adult mass.  For example, making use of the observation that body fat in
mammals scales with overall body size according to $M_{f}=f_{0}M^{\gamma}$
and assuming that once this mass is fully digested the organism begins to
starve, this would imply that $\epsilon=1-f_{0}M^{\gamma}/M$.  Using this
criterion in Eq.~\ref{mt}, the time scale for starvation is given by
\begin{equation}
\label{eq:sigma}
t_{\sigma}=-\frac{E_{m}\ln\left(\epsilon\right)}{B_{m}}.
\end{equation}
The starvation rate is then $\sigma=1/t_{\sigma}$, which scales with
replete-state mass as $1/\ln\left(1-f_{0}M^{\gamma}/M\right)$.  An important
feature is that $\sigma$ does not have a simple scaling dependence on
$\lambda$ (Eq.~\ref{lambda}), which is important for the dynamics that we
later discuss.

@@

The rate of recovery $\rho = 1/t_\rho$ requires that an organism accrues
sufficient tissue to transition from the starving \sid{starving vs.\ hungry}
state to the full state.  We again use the balance given in Eq.~\ref{balance}
to find the timescale to return to the replete-state mass from a given
reduced starvation mass.  From the solution to Eq.~\ref{balance}
\begin{eqnarray}
m\left(t\right)\!=\!\left(\!\frac{B_{0}}{B_{m}}\!\right)^{1/(\eta-1)}\!\!
\left[1\!-\!\left(1\!-\!\frac{B_{m}}{B_{0}}m_{0}^{1\!-\!\eta}\right)e^{-b\left(1\!-\!\eta\right)t}\right]^{1/\left(1-\eta\right)}%\left(\frac{a}{b}\right)^{1/\left(1-\eta\right)}\,,
\end{eqnarray}
we require the timescale, $t_{\rho}=t_{2}-t_{1}$, which is the time it takes
to go from $m\left(t_{1}\right)=\epsilon M$ to $m\left(t_{2}\right)=M$, or
\begin{eqnarray}
t_{\rho}=\frac{E_{m}\left\{\ln \left[1\!-\!\frac{B_{0}}{B_{m}}\left(M
  \right)^{1\!-\!\eta }\right]
-\ln \left[1\!-\!\frac{B_{0}}{B_{m}}\left(\epsilon M \right)^{1\!-\!\eta }\right]\right\}}{(\eta -1) B_{m}}.
\end{eqnarray}
%\left(\frac{a}{b}\right)^{\frac{1}{\eta -1}}
%where $c=(a/b)^{1/(\eta-1)}$.
Although these rate equations are general, here we focus on parameterizations
for terrestrial-bound endotherms, specifically mammals, which range from a
minimum of $M\approx1$ gram (the Etruscan shrew \emph{Suncus etruscus}) to a
maximum of $M\approx10^7$ grams (the late Eocene to early Miocene
Indricotheriinae).  Investigating other classes of organisms would simply
involve altering the metabolic exponents and scalings associate with
$\epsilon$. Moreover, we emphasize that our allometric equations describe
mean relationships, and do not account for the (sometimes considerable)
variance associated with individual species.
\\
%, though this is beyond the current scope of our analysis. %, and we henceforth focus our examination on mammalian species.

@@

%Minimal mass bound
A lower bound on mammalian body size is given by $\epsilon=1$, where mammals
have no metabolic reserves and immediately starve; this occurs at a size of
\sid{$M=value$}.  This calculation \sid{what calculation?} gives an extreme
limit on size but does not account for the subtleties of starvation dynamics
that may limit body size. The NSM correctly predicts that species with
smaller masses have larger steady-state population densities.  However we
observe that there is a sharp change in the mass dependence of both the
steady-state densities and $\sigma/\rho$ at $M \approx 0.3$ grams
(Fig.~\ref{fig:mass}a,b).  The dependence of the rates of starvation and
recovery explain this phenomenon.  As the mass decreases, the rate of
starvation increases, while the rate of recovery declines super-exponentially
\sid{how do we know this?}.  This decline in $\rho$ occurs when the
percentage of body fat is
$1 - 1/\left[\big(B_{0}/B_{m}\big)^{1/\left(\eta-1\right)}M\right] \approx
2\%$,
whereupon consumers have no eligible route \sid{what does this mean?} out of
starvation.  Compellingly, this dynamic bound determined by the rate of
energetic recovery is close to the minimum observed mammalian body size of
ca. 1.3-2.5 grams (Fig.~\ref{fig:mass}b,c), a range that occurs as the
recovery rate begins its decline.  In addition to known transport
limitations~\cite{West:2002ud}, we suggest that an additional constraint of
lower body size stems from the dynamics of starvation.  This work \sid{which
  work?} mirrors other efforts where coincident limitations seem to limit the
smallest possibilities for life within a particular class or
organisms~\cite{kempes}.

@@

The observed switch in susceptibility as a function of $\chi$ at
$M_{\rm opt}\approx10^6$ thus serves as an attractor, where over evolutionary
times the NSM predicts organismal mass to increase if $M<M_{\rm opt}$ and
decrease if $M>M_{\rm opt}$.  Moreover, $M_{\rm opt}$, which is entirely
determined by the population-level consequences of energetic constraints, is
within an order of magnitude of the mass observed in the North American
mammalian fossil record~\cite{Alroy:1998p1594} and also the mass predicted
from an evolutionary model of body size evolution~\cite{Clauset:2009fh}.
While the state of the environment, as well as the competitive landscape,
will determine whether specific body sizes are selected for or
against~\cite{Saarinen:2014br}, we suggest that the starvation dynamic
proposed here may provide the driving mechanism for the evolution of larger
body size among terrestrial mammals.
%and we suggest that the starvation dynamics described here act in concert with these other factors is the only driver of body size evolution.

@@

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{Growth-trajectory-diagram.pdf}
\caption{ % 58 words
}
\label{fig:growth}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{fig_GvS.pdf}
\caption{ % 58 words
}
\label{fig:gvs}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{fig_FixedPoint.pdf}
\caption{ % 58 words
}
\label{fig:fp}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{fig_DataHopf.pdf}
\caption{ % 58 words
}
\label{fig:hopf}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{fig_ExtinctionAllometric.pdf}
\caption{ % 58 words
}
\label{fig:ext}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.45\textwidth]{fig_FPAllometric.pdf}
\caption{ % 58 words
}
\label{fig:mass}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{fig_Invasion.pdf}
\caption{ % 58 words
}
\label{fig:invasion}
\end{figure*}

>>>>

\noindent {\bf Nutritional-state-structured model (NSM)}\\
We begin by defining a minimal Nutritional-State-structured population Model
(NSM), where the consumer population is divided into two energetic states:
(a) an energetically replete (full) state $F$, where the consumer reproduces
at a constant rate $\lambda$ and has no mortality risk, and (b) an
energetically deficient (hungry) state $H$, where the consumer does not
reproduce but dies at rate $\mu$.  The underlying resource $R$ evolves by
logistic growth with an intrinsic growth rate $\alpha$ and a carrying
capacity equal to one.  Consumers transition from the full state $F$ to the
hungry state $H$ at a rate $\sigma$---the starvation rate---and also in
proportion to the absence of resources $(1-R)$.  Conversely, consumers
recover from state $H$ to state $F$ at rate $\rho$ and in proportion to $R$.
Resources are also eaten by the consumers---at rate $\rho$ by hungry
consumers and at rate $\beta<\rho$ by full consumers.  This inequality
accounts for hungry consumers requiring more resources to rebuild body
weight.

@@

Nearly all of the rates described in the NSM are to some extent governed by
consumer metabolism, which can be used to describe a variety of organismal
features~\cite{}\sid{Ref}.  The scaling relation between an organism's
metabolic rate $B$ and its body size at reproductive maturity $M$ is well
documented~\cite{West:2002it} and scales as $B = B_0 M^\eta$, where $\eta$ is
the scaling exponent, generally assumed to vary around $2/3$ or $3/4$ for
metazoans~\cite{}\sid{Ref}, and has taxonomic shifts for unicellular species
between $\eta\approx 1$ in eukaryotes and $\eta\approx 1.76$ in bacteria
\cite{DeLong:2010dy,Kempes:2012hy}. Several efforts have shown how a
partitioning of this metabolic rate between growth and maintenance purposes
can be used to derive a general equation for the growth trajectories and
growth rates of organisms ranging from bacteria to metazoans
\cite{Kempes:2012hy,more}\sid{fix Ref}. More specifically, the interspecific
\sid{what does interspecific mean?}  trends in growth rate can be
approximated by
\begin{eqnarray}
\lambda = \lambda_0 M^{\eta-1}.
\label{lambda}
\end{eqnarray}
This relation is derived from the simple balance condition
\begin{eqnarray}
\label{balance}
B_{0}m^{\eta}=E_{m}\frac{dm}{dt}+B_{m}m\,,
\end{eqnarray}
where $E_{m}$ is the energy needed to synthesize a unit of mass, $B_{m}$ is
the metabolic rate to support an existing unit of mass, and $m$ is the mass
at any point in development.  It is useful to explicitly write this balance
because it can also be modified to understand the timescales of starvation as
well as recovery from starvation as we show below.

To determine the starvation rate, $\sigma$, we are interested in the time
required for an organism to go from a mature adult that reproduces at rate
$\lambda$ (henceforth we term this state as the ``replete'' state), to a
reduced-mass hungry state where reproduction is impossible.  This transition
time can be inferred from the energy balance in Eq.~\ref{balance}, where we
make the basic assumption that an organism must meet its maintenance
requirements using the digestion of existing mass as the sole energy source.
This assumption implies the simple metabolic balance \sid{don't understand
  this equation}
\begin{eqnarray}
\frac{dm}{dt}E_{m}^{\prime}=-B_{m}m
\end{eqnarray}
where $E_{m}^{\prime}$ is the amount of energy stored in a unit of existing
body mass which differs from $E_{m}$~\cite{}\sid{Ref}, the energy required to
synthesis a unit of biomass. Given the replete mass, $M$, of an organism, the
above energy balance prescribes the mass trajectory of a non-consuming
organism:
\begin{eqnarray}
\label{mt}
m\left(t\right)=Me^{-B_{m}t/E_{m}^{\prime}}.
\end{eqnarray}
Since only certain tissues can be digested for energy (for example the brain
cannot be degraded to fuel metabolism), we define the rate for starvation and
death by the timescales required to reach specific fractions of the
replete-state mass.  We define $m_{\rm starve}=\epsilon M$, where
$\epsilon<1$ is the fraction of replete-state mass where reproduction ceases.
This fraction will be modified if tissue composition systematically scales
with adult mass.  For example, making use of the observation that body fat in
mammals scales with overall body size according to $M_{f}=f_{0}M^{\gamma}$
and assuming that once this mass is fully digested the organism starves, this
would imply that $\epsilon=1-f_{0}M^{\gamma}/M$.  Using this criterion in
Eq.~\ref{mt}, the time scale for starvation is given by
\begin{equation}
\label{eq:sigma}
t_{\sigma}=-\frac{E_{m}\ln\left(\epsilon\right)}{B_{m}}.
\end{equation}
The starvation rate is then $\sigma=1/t_{\sigma}$, which scales with
replete-state mass as $1/\ln\left(1-f_{0}M^{\gamma}/M\right)$.  An important
feature is that $\sigma$ does not have a simple scaling dependence on
$\lambda$ (Eq.~\ref{lambda}), which is important for the dynamics that we
later discuss.

@@

The rate of recovery $\rho = 1/t_\rho$ requires that an organism accrues
sufficient tissue to transition from the hungry to the full state.  We again
use the balance given in Eq.~\ref{balance} to find the timescale for an
organism to return to the replete-state mass from a given reduced
hungry-state mass.  From the solution to Eq.~\ref{balance}
\begin{eqnarray}
m\left(t\right)\!=\!\left(\!\frac{B_{0}}{B_{m}}\!\right)^{1/(\eta-1)}\!\!
\left[1\!-\!\left(1\!-\!\frac{B_{m}}{B_{0}}m_{0}^{1\!-\!\eta}\right)e^{-b\left(1\!-\!\eta\right)t}\right]^{1/\left(1-\eta\right)}%\left(\frac{a}{b}\right)^{1/\left(1-\eta\right)}\,,
\end{eqnarray}
we require the timescale, $t_{\rho}=t_{2}-t_{1}$, which is the time it takes
to go from $m\left(t_{1}\right)=\epsilon M$ to $m\left(t_{2}\right)=M$, or
\begin{eqnarray}
t_{\rho}=\frac{E_{m}\left\{\ln \left[1\!-\!\frac{B_{0}}{B_{m}}\left(M
  \right)^{1\!-\!\eta }\right]
-\ln \left[1\!-\!\frac{B_{0}}{B_{m}}\left(\epsilon M \right)^{1\!-\!\eta }\right]\right\}}{(\eta -1) B_{m}}.
\end{eqnarray}
%\left(\frac{a}{b}\right)^{\frac{1}{\eta -1}}
%where $c=(a/b)^{1/(\eta-1)}$.
Although these rate equations are general, here we focus on parameterizations
for terrestrial-bound endotherms, specifically mammals, which range from a
minimum of $M\approx1$ gram (the Etruscan shrew \emph{Suncus etruscus}) to a
maximum of $M\approx10^7$ grams (the late Eocene to early Miocene
Indricotheriinae).  Investigating other classes of organisms would simply
involve altering the metabolic exponents and scalings associate with
$\epsilon$. Moreover, we emphasize that our allometric equations describe
mean relationships, and do not account for the (sometimes considerable)
variance associated with individual species.
\\
%, though this is beyond the current scope of our analysis. %, and we henceforth focus our examination on mammalian species.

@@

%Minimal mass bound
A lower bound on mammalian body size is given by $\epsilon=1$, where mammals
have no metabolic reserves and immediately starve; this occurs at a size of
\sid{$M=value$}.  This calculation \sid{what calculation?} gives an extreme
limit on size but does not account for the subtleties of starvation dynamics
that may limit body size. The NSM correctly predicts that species with
smaller masses have larger steady-state population densities.  However we
observe that there is a sharp change in the mass dependence of both the
steady-state densities and $\sigma/\rho$ at $M \approx 0.3$ grams
(Fig.~\ref{fig:mass}a,b).  The dependence of the rates of starvation and
recovery explain this phenomenon.  As the mass decreases, the rate of
starvation increases, while the rate of recovery declines super-exponentially
\sid{how do we know this?}.  This decline in $\rho$ occurs when the
percentage of body fat is
$1 - 1/\left[\big(B_{0}/B_{m}\big)^{1/\left(\eta-1\right)}M\right] \approx
2\%$,
whereupon consumers have no eligible route \sid{what does this mean?} that avoids
starvation.  Compellingly, this dynamic bound determined by the rate of
energetic recovery is close to the minimum observed mammalian body size of
ca. 1.3-2.5 grams (Fig.~\ref{fig:mass}b,c), a range that occurs as the
recovery rate begins its decline.  In addition to known transport
limitations~\cite{West:2002ud}, we suggest that an additional constraint of
lower body size stems from the dynamics of starvation.  This work \sid{which
  work?} mirrors other efforts where coincident limitations seem to limit the
smallest possibilities for life within a particular class or
organisms~\cite{kempes}.

@@

The observed switch in susceptibility as a function of $\chi$ at
$M_{\rm opt}\approx10^6$ thus serves as an attractor, where over evolutionary
times the NSM predicts organismal mass to increase if $M<M_{\rm opt}$ and
decrease if $M>M_{\rm opt}$.  Moreover, $M_{\rm opt}$, which is entirely
determined by the population-level consequences of energetic constraints, is
within an order of magnitude of the mass observed in the North American
mammalian fossil record~\cite{Alroy:1998p1594} and also the mass predicted
from an evolutionary model of body size evolution~\cite{Clauset:2009fh}.
While the state of the environment, as well as the competitive landscape,
will determine whether specific body sizes are selected for or
against~\cite{Saarinen:2014br}, we suggest that the starvation dynamics
proposed here may provide the driving mechanism for the evolution of larger
body size among terrestrial mammals.
%and we suggest that the starvation dynamics described here act in concert with these other factors is the only driver of body size evolution.

@@

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{Growth-trajectory-diagram.pdf}
\caption{
The growth trajectory over absolute time of an individual organism as a function of body mass.
Initial growth follows the red trajectory to an energetically replete adult mass $M$.
Starvation follows the concave blue trajectory to $m_{\rm starve} < M$, whereas recovery follows the convex growth trajectory from $m_\sigma$ to $M$.
}
\label{fig:growth}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{fig_GvS.pdf}
\caption{
Allometrically constrained starvation rate $\sigma$ (red) vs. reproductive rate $\lambda$ (yellow) as a function of mass $M$.
The rate of starvation is greater than the rate of reproduction for all realized terrestrial endotherm body sizes.
}
\label{fig:gvs}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{fig_FixedPoint.pdf}
\caption{
The transcritical (TC; dashed line) and Hopf bifurcation (solid line) as a function of the starvation rate $\sigma$ and recovery rate $\rho$.
These bifurcation conditions separate parameter space into infeasible, cyclic, and steady state dynamic regimes.
The color gradient shows the steady state densities for (A) the resource $R^*$ and the (B) energetically replete consumers $F^*$, with warm colors denoting higher densities and cool colors denoting lower densities.
Steady state densities for the energetically deficient consumers $H^*$ are not shown because they closely mirror those for $F^*$.
}
\label{fig:fp}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{fig_DataHopf.pdf}
\caption{
Transcritical (TC; vertical lines) and Hopf bifurcations (curved lines) with allometrically determined starvation $\sigma$ and recovery $\rho$ rates as a function of minimum and maximum mammalian body sizes: $1$ gram (blue) and $10^7$ grams (orange), respectively.
Replicates show the influence of variation (20\% around the mean) on allometric parameters, which influences both the energetic rates as well as the position of the TC and Hopf bifurcations.
}
\label{fig:hopf}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{fig_ExtinctionAllometric.pdf}
\caption{
The probability of extinction for 1000 consumer population trajectories as a function of $\sigma/\rho$ within initial densities chosen to as $A(F^*,H^*,R^*)$, with $A$ a random
variable that is uniformly distribution in $[0,2]$.
Extinction is defined as the population trajectory going below
$0.2~\times$ the allometrically constrained steady state for all times
between $10^2$ and $\leq 10^6$.
The values above the extinction plot are the allometrically constrained $\sigma/\rho$ with 20\% variation around the mean.
}
\label{fig:ext}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.45\textwidth]{fig_FPAllometric.pdf}
\caption{
(\emph{A}) Consumer steady states as a function of body size, showing both energetically deficient and replete consumer states ($H^*$ and $F^*$, respectively). 
Energetic rates as a function of body size, with the ratio $\sigma/\rho$ (\emph{B}) and both $\sigma$ (red) and $\rho$ (purple; \emph{C}) drawn separately with 20\% variation around the mean.
Steady state densities decline sharply at $M=M_{\rm min}$ due to the super-exponential decrease in the rate of recovery.
The minimum body size observed for mammals (the Etruscan shrew) is denoted by the orange shaded region at values marking the initial decline of the recovery rate.
}
\label{fig:mass}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{fig_Invasion.pdf}
\caption{
Invasion feasibility for organisms with a proportional change in mass $\chi$ against a population with a resident body mass $M$.
The blue region denotes which values of $\chi$ result in successful invasion.
The red region denotes which values of $\chi$ result in a mass that is below the starvation threshold and is thus infeasible.
}
\label{fig:invasion}
\end{figure*}

### 2016-07-29 Manuscript/5.tex
Kempes edits

Nearly all of the rates described in the NSM are to some extent governed by
consumer metabolism, which can be used to describe a variety of organismal
features~\cite{}\sid{Ref}.  The scaling relation between an organism's
metabolic rate $B$ and its body size at reproductive maturity $M$ is well
documented~\cite{West:2002it} and scales as $B = B_0 M^\eta$, where $\eta$ is
the scaling exponent, generally assumed to vary around $2/3$ or $3/4$ for
metazoans~\cite{}\sid{Ref}, and has taxonomic shifts for unicellular species
between $\eta\approx 1$ in eukaryotes and $\eta\approx 1.76$ in bacteria
\cite{DeLong:2010dy,Kempes:2012hy}. Several efforts have shown how a
partitioning of this metabolic rate between growth and maintenance purposes
can be used to derive a general equation for the growth trajectories and
growth rates of organisms ranging from bacteria to metazoans
\cite{Kempes:2012hy,more}\sid{fix Ref}. More specifically, the interspecific
\sid{what does interspecific mean?}  trends in growth rate can be
approximated by
\begin{eqnarray}
\lambda = \lambda_0 M^{\eta-1}.
\label{lambda}
\end{eqnarray}
This relation is derived from the simple balance condition
\begin{eqnarray}
\label{balance}
B_{0}m^{\eta}=E_{m}\frac{dm}{dt}+B_{m}m\,,
\end{eqnarray}
where $E_{m}$ is the energy needed to synthesize a unit of mass, $B_{m}$ is
the metabolic rate to support an existing unit of mass, and $m$ is the mass
at any point in development.  It is useful to explicitly write this balance
because it can also be modified to understand the timescales of starvation as
well as recovery from starvation as we show below.

To determine the starvation rate, $\sigma$, we are interested in the time
required for an organism to go from a mature adult that reproduces at rate
$\lambda$ (henceforth we term this state as the ``replete'' state), to a
reduced-mass hungry state where reproduction is impossible.  This transition
time can be inferred from the energy balance in Eq.~\ref{balance}, where we
make the basic assumption that an organism must meet its maintenance
requirements using the digestion of existing mass as the sole energy source.
This assumption implies the simple metabolic balance \sid{don't understand
  this equation}
\begin{eqnarray}
\frac{dm}{dt}E_{m}^{\prime}=-B_{m}m
\end{eqnarray}
where $E_{m}^{\prime}$ is the amount of energy stored in a unit of existing
body mass which differs from $E_{m}$~\cite{}\sid{Ref}, the energy required to
synthesis a unit of biomass. Given the replete mass, $M$, of an organism, the
above energy balance prescribes the mass trajectory of a non-consuming
organism:
\begin{eqnarray}
\label{mt}
m\left(t\right)=Me^{-B_{m}t/E_{m}^{\prime}}.
\end{eqnarray}
Since only certain tissues can be digested for energy (for example the brain
cannot be degraded to fuel metabolism), we define the rate for starvation and
death by the timescales required to reach specific fractions of the
replete-state mass.  We define $m_{\rm starve}=\epsilon M$, where
$\epsilon<1$ is the fraction of replete-state mass where reproduction ceases.
This fraction will be modified if tissue composition systematically scales
with adult mass.  For example, making use of the observation that body fat in
mammals scales with overall body size according to $M_{f}=f_{0}M^{\gamma}$
and assuming that once this mass is fully digested the organism starves, this
would imply that $\epsilon=1-f_{0}M^{\gamma}/M$.  Using this criterion in
Eq.~\ref{mt}, the time scale for starvation is given by
\begin{equation}
\label{eq:sigma}
t_{\sigma}=-\frac{E_{m}\ln\left(\epsilon\right)}{B_{m}}.
\end{equation}
The starvation rate is then $\sigma=1/t_{\sigma}$, which scales with
replete-state mass as $1/\ln\left(1-f_{0}M^{\gamma}/M\right)$.  An important
feature is that $\sigma$ does not have a simple scaling dependence on
$\lambda$ (Eq.~\ref{lambda}), which is important for the dynamics that we
later discuss.

>>>>

Nearly all of the rates described in the NSM are to some extent governed by
consumer metabolism, which can be used to describe a variety of organismal
features~\cite{}\sid{Ref}.  The scaling relation between an organism's
metabolic rate $B$ and its body size at reproductive maturity $M$ is well
documented~\cite{West:2002it} and scales as $B = B_0 M^\eta$, where $\eta$ is
the scaling exponent, generally assumed to vary around $2/3$ or $3/4$ for
metazoans~\cite{}\sid{Ref}, and has taxonomic shifts for unicellular species
between $\eta\approx 1$ in eukaryotes and $\eta\approx 1.76$ in bacteria
\cite{DeLong:2010dy,Kempes:2012hy}. Several efforts have shown how a
partitioning of this metabolic rate between growth and maintenance purposes
can be used to derive a general equation for the growth trajectories and
growth rates of organisms ranging from bacteria to metazoans
\cite{Kempes:2012hy,more}\sid{fix Ref}. More specifically, the cross-species trends in growth rate can be
approximated by
\begin{eqnarray}
\lambda = \lambda_0 M^{\eta-1}.
\label{lambda}
\end{eqnarray}
This relation is derived from the simple balance condition
\begin{eqnarray}
\label{balance}
B_{0}m^{\eta}=E_{m}\frac{dm}{dt}+B_{m}m\,,
\end{eqnarray}
where $E_{m}$ is the energy needed to synthesize a unit of mass, $B_{m}$ is
the metabolic rate to support an existing unit of mass, and $m$ is the mass
at any point in development.  It is useful to explicitly write this balance
because it can also be modified to understand the timescales of starvation as
well as recovery from starvation as we show below.

To determine the starvation rate, $\sigma$, we are interested in the time
required for an organism to go from a mature adult that reproduces at rate
$\lambda$ (henceforth we term this state as the ``replete'' state), to a
reduced-mass hungry state where reproduction is impossible.  This transition
time can be inferred from the energy balance in Eq.~\ref{balance}, where we
make the basic assumption that an organism must meet its maintenance
requirements using the digestion of existing mass as the sole energy source.
This assumption implies a simple metabolic balance where the left hand side of Equation \ref{balance} becomes zero and we have \begin{eqnarray}
\frac{dm}{dt}E_{m}^{\prime}=-B_{m}m
\end{eqnarray}
where $E_{m}^{\prime}$ is the amount of energy stored in a unit of existing
body mass which differs from $E_{m}$~\cite{}\sid{Ref}, the energy required to
synthesis a unit of biomass. Given the replete mass, $M$, of an organism, the
above energy balance prescribes the mass trajectory of a non-consuming
organism:
\begin{eqnarray}
\label{mt}
m\left(t\right)=Me^{-B_{m}t/E_{m}^{\prime}}.
\end{eqnarray}
Since only certain tissues can be digested for energy (for example the brain
cannot be degraded to fuel metabolism), we define the rate for starvation and
death by the timescales required to reach specific fractions of the
replete-state mass.  We define $m_{\rm starve}=\epsilon M$, where
$\epsilon<1$ is the fraction of replete-state mass where reproduction ceases.
This fraction will be modified if tissue composition systematically scales
with adult mass.  For example, making use of the observation that body fat in
mammals scales with overall body size according to $M_{f}=f_{0}M^{\gamma}$
and assuming that once this mass is fully digested the organism starves, this
would imply that $\epsilon=1-f_{0}M^{\gamma}/M$.  Using this criterion in
Eq.~\ref{mt}, the time scale for starvation is given by
\begin{equation}
\label{eq:sigma}
t_{\sigma}=-\frac{E_{m}\ln\left(\epsilon\right)}{B_{m}}.
\end{equation}
The starvation rate is then $\sigma=1/t_{\sigma}$, which scales with
replete-state mass as $1/\ln\left(1-f_{0}M^{\gamma}/M\right)$.  An important
feature is that $\sigma$ does not have a simple scaling dependence on
$\lambda$ (Eq.~\ref{lambda}), which is important for the dynamics that we
later discuss.
